Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-25 Manuscript/SupportingInfo.tex
vsn 4

Here we provide additional supporting information referenced in the main
document, which include additional analyses, and simulation results
across a wider set of input conditions. As before, simulations were
conducted on six different tree sizes (\(n=2^5, 2^6, \cdots, 2^{10}\)),
and with differing levels of phylogenetic signal
(\(\lambda=0.0, 0.5, \cdots, 1.0\)). We generated 100 random trees for
each intersection of tree size and \(\lambda\). For each \(\lambda\)
within each tree size, continuous traits were then simulated on each
phylogeny under a BM model of evolution. For each set of 100 trees we
measured the mean values of \(\hat{\lambda}\) and \(\kappa\), their
standard deviation, and calculated the Shapiro-Wilk \(W\) statistic as a
departure from normality (symmetry). For the latter, a value of \(1.0\)
indicates normally distributed values, while departures from \(1.0\)
indicate skewness.

@@

Results from the simulations on pure-birth trees revealed that both
\(Z_{\lambda}\) and \(Z_K\) increased with increasing input phylogenetic
signal, but \(\hat{\lambda}\) was strongly affected by tree size,
whereas \(Z_K\) was m1re consistent (Fig. S4). Additionally, \(Z_K\)
increased more linearly with increasing levels of phylogenetic signal,
and its standard deviation across input signal was more even across tree
sizes, implying more consistent precision. This latter observation is
seen in the upper four panels of Fig. S1, where the dispersion of
simulation results (gray dots) is far more even at each input level of
phylogenetic signal for \(Z_K\) than it is for \(Z_{\lambda}\).

@@

\textbf{Figure S1}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion.

@@

For \emph{K}, mean values increased with increasing phylogenetic signal,
though as was found with pure-birth trees, the increase was nonlinear
(Fig. S3 black line). Likewise, variation increased with increasing
phylogenetic signal (Fig. S3 red line), though the distribution of
\emph{K} was more normally distributed throughout its range, and across
different tree sizes, as compared with \(\hat{\lambda}\), thought there
was some slight skew for high input levels of phylogenetic signal on
large phylogenies (Fig. S3 blue line). \hfill\break

@@

\textbf{Figure S2}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pectinate trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

\textbf{Figure S3}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on pectinate trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

\textbf{Figure S4}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion.

@@

\textbf{Figure S5}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on balanced trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

\textbf{Figure S6}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on balanced trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

\textbf{Figure S7}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion.

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 50\%
collapsed nodes, following the procedures of (1). Results from
simulations on phylogenies containing polytomies mirrored those found on
fully resolved trees. For \(\hat{\lambda}\), the mean value increased
with increasing input signal, but was negatively biased, and was less
than the input value across most of its range (Fig. S8: black line).
Additionally, the precision of \(\hat{\lambda}\) varied with differing
input levels, with the greatest variation found at intermediate values
of \(\lambda\) (Fig. S8 red line). Finally, the distribution of
\(\hat{\lambda}\) was not normal, and became more skewed at more extreme
values of \(\lambda\) (Fig. S8 blue line). \hfill\break

@@

\textbf{Figure S8}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pure-birth trees containing polytomies.
Gray line signifies the 1:1 line where the input value matches the
estimate.

@@

\textbf{Figure S9}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on pure-birth trees containing polytomies.
Gray line signifies the 1:1 line where the input value matches the
estimate.

@@

\textbf{Figure S10}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_K\) to increasing strength of Brownian motion.

@@

We also investigated the effect of incorporating \(\lambda\) in PGLS
analyses (phylogenetic regression and ANOVA). Here, patterns for
\(\hat\lambda\), \emph{K}, \(Z_\lambda\) and \(Z_\kappa\) were virtually
identical to those found previously. In essence, \(\hat\lambda\) was not
suitable as a test statistic representing phylogenetic signal due to all
properties previously shown (Figs. S11 \& S12), \emph{K} was more
appropriate, but associated non-linearly with signal strength (Figs. S13
\& 14), while \(Z_\kappa\) was found to be a superior effect size as
compared with \(Z_\lambda\) (Figs S15 \& S16). \hfill\break

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic regression
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

@@

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

@@

\textbf{Figure S19}. Power curves for phylogenetic regression
(incorporating \(\lambda\)). Type I error is found on the far-left of
the plot.

@@

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating
\(\lambda\)). Type I error is found on the far-left of the plot.

>>>>

Here we provide additional supporting information referenced in the main
document, which include additional analyses, and simulation results
across a wider set of input conditions. As before, simulations were
conducted on six different tree sizes (\(n=2^5, 2^6, \cdots, 2^{10}\)),
and with differing levels of phylogenetic signal
(\(\lambda=0.0, 0.5, \cdots, 1.0\)). We generated 50 random trees for
each intersection of tree size and \(\lambda\). For each \(\lambda\)
within each tree size, continuous traits were then simulated on each
phylogeny under a BM model of evolution. For each set of 50 trees we
measured the mean values of \(\hat{\lambda}\) and \(K\), their standard
deviation, and calculated the Shapiro-Wilk \(W\) statistic as a
departure from normality (symmetry). For the latter, a value of \(1.0\)
indicates normally distributed values, while departures from \(1.0\)
indicate skewness.

@@

Results from the simulations on pure-birth trees revealed that both
\(Z_{\lambda}\) and \(Z_K\) increased with increasing input phylogenetic
signal, but \(\hat{\lambda}\) was strongly affected by tree size,
whereas \(Z_K\) was more consistent (Fig. S4). Additionally, \(Z_K\)
increased more linearly with increasing levels of phylogenetic signal,
and its standard deviation across input signal was more even across tree
sizes, implying more consistent precision. This latter observation is
seen in the upper four panels of Fig. S1, where the dispersion of
simulation results (gray dots) is far more even at each input level of
phylogenetic signal for \(Z_K\) than it is for \(Z_{\lambda}\).

@@

\textbf{Figure S1}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion for pure-birth trees.

@@

For \emph{K}, mean values increased with increasing phylogenetic signal,
though as was found with pure-birth trees, the increase was nonlinear
(Fig. S3 black line). Likewise, variation increased with increasing
phylogenetic signal (Fig. S3 red line), though the distribution of
\emph{K} was more normally distributed throughout its range, and across
different tree sizes, as compared with \(\hat{\lambda}\), though there
was some slight skew for high input levels of phylogenetic signal on
large phylogenies (Fig. S3 blue line). \hfill\break

@@

\textbf{Figure S2}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pectinate trees. Gray line signifies the
1:1 line where the input value matches the estimate.At each input level,
the dark black line represents the empirically derived expected value
(mean) of \(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

@@

\textbf{Figure S3}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on pectinate trees. At each input level, the
dark black line represents the empirically derived expected value (mean)
of \textit{K}, the red line is the standard deviation of \textit{K}, and
the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\)
signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S4}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion for pectinate trees.

@@

\textbf{Figure S5}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on balanced trees. Gray line signifies the
1:1 line where the input value matches the estimate. At each input
level, the dark black line represents the empirically derived expected
value (mean) of \(\hat\lambda\), the red line is the standard deviation
of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

@@

\textbf{Figure S6}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on balanced trees. At each input level, the
dark black line represents the empirically derived expected value (mean)
of \textit{K}, the red line is the standard deviation of \textit{K}, and
the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\)
signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S7}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion for balanced trees.

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 20\%
collapsed nodes, following the procedures of (1). Results from
simulations on phylogenies containing polytomies mirrored those found on
fully resolved trees. For \(\hat{\lambda}\), the mean value increased
with increasing input signal, but was negatively biased, and was less
than the input value across most of its range (Fig. S8: black line).
Additionally, the precision of \(\hat{\lambda}\) varied with differing
input levels, with the greatest variation found at intermediate values
of \(\lambda\) (Fig. S8 red line). Finally, the distribution of
\(\hat{\lambda}\) was not normal, and became more skewed at more extreme
values of \(\lambda\) (Fig. S8 blue line). \hfill\break

@@

\textbf{Figure S8}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pure-birth trees containing polytomies.
Gray line signifies the 1:1 line where the input value matches the
estimate. At each input level, the dark black line represents the
empirically derived expected value (mean) of \(\hat\lambda\), the red
line is the standard deviation of \(\hat\lambda\), and the blue line is
Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies
normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S9}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion on pure-birth trees containing polytomies.
At each input level, the dark black line represents the empirically
derived expected value (mean) of \textit{K}, the red line is the
standard deviation of \textit{K}, and the blue line is Shapiro Wilks
statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\)
represent skewed distributions).

@@

\textbf{Figure S10}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_K\) to increasing strength of Brownian motion for pure-birth trees
containing polytomies.

@@

We also investigated the effect of incorporating \(\lambda\) in PGLS
analyses (phylogenetic regression and ANOVA). Here, patterns for
\(\hat\lambda\), \emph{K}, \(Z_\lambda\) and \(Z_K\) were virtually
identical to those found previously. In essence, \(\hat\lambda\) was not
suitable as a test statistic representing phylogenetic signal due to all
properties previously shown (Figs. S11 \& S12), \emph{K} was more
appropriate, but associated non-linearly with signal strength (Figs. S13
\& 14), while \(Z_K\) was found to be a superior effect size as compared
with \(Z_\lambda\) (Figs S15 \& S16). \hfill\break

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate. At each input level, the dark black line
represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

@@

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate. At each input level, the dark black line
represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

@@

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). At each input level, the dark black line represents the
empirically derived expected value (mean) of \textit{K}, the red line is
the standard deviation of \textit{K}, and the blue line is Shapiro Wilks
statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\)
represent skewed distributions).

@@

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). At each input level, the dark black line represents the
empirically derived expected value (mean) of \textit{K}, the red line is
the standard deviation of \textit{K}, and the blue line is Shapiro Wilks
statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\)
represent skewed distributions).

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic regression
(incorporating \(\lambda\)). The red line represents the estimated
slope, which almost completely overlaps the gray 1:1 line.

@@

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA
(incorporating \(\lambda\)). The red line represents the estimated
slope, which almost completely overlaps the gray 1:1 line.

@@

\textbf{Figure S19}. Power curves for phylogenetic regression
(incorporating \(\lambda\)). Type I error is found on the far-left of
the plot (effect = 0).

@@

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating
\(\lambda\)). Type I error is found on the far-left of the plot (effect
= 0).
